\newpage{}

# Introduction

R [@R] has the ability to display mathematical symbols and equations in graphics using the "plotmath" feature [@plotmath], interpreting everything within a call to `expression()` as a mathematical equation.

```{r expressionPlot, fig.cap="A plot with axis labels made using \\texttt{expression()}.", include = TRUE, eval = TRUE, fig.asp = 1}
x <- 1:5
y <- x ^ 2 / 2
expr <- expression(frac(mu ^ 2, 2))
plot(x, y, xlab = expression(mu), ylab = "", yaxt = "n", type = "l")
axis(2, las = 1)
mtext(expr, side = 2, line = 3, las = 1)
```

This provides us with most of the symbols used for equations, such as brackets and fractions, and formats them in a layout resembling \TeX{} [@knuth-tex], but it is limited in its fonts. Compare the y-axis title above with how it looks when created by \LaTeX{} in figure \ref{muOver2}.

\begin{figure}\label{muOver2}
\begin{equation*}
\dfrac{\mu^2}{2}
\end{equation*}
\caption{The y-axis title from above, created with \LaTeX}
\end{figure}

The difference is stark and @murrell-dvir describes several approaches in R, and their limitations, which can get us closer to the \LaTeX{} result as motivation for the `dvir` package. The approaches are replicated here.

The `extrafont` package [@pkg-extrafont] with the `fontcm` package extension [@pkg-fontcm] works for this particular example but not if we wanted to include other \TeX{} symbols.

```{r, eval = FALSE}
library(grid)
library(extrafont)
font_install('fontcm')
loadfonts("pdf")
pdf("extrafont.pdf", width = 1, height = 1)
grid.text(expr, gp = gpar(fontfamily = "CM Roman"))
dev.off()
embed_fonts("extrafont.pdf", outfile = "extrafont-embed.pdf")
```

```{r, fig.cap = "The axis title using the \\texttt{extrafont} package", echo = FALSE, out.width = "40%"}
include_graphics("../Figures/extrafont-embed.pdf")
```

```{r, eval = FALSE}
expr2 <- expression(bgroup("(", frac(x - mu, sigma), ")"))
pdf("extrafont-2.pdf", width = 1, height = 1)
grid.text(expr2, gp = gpar(fontfamily = "CM Roman"))
dev.off()
embed_fonts("extrafont-2.pdf", outfile = "extrafont-2-embed.pdf")
```

```{r bracExampleExtra, fig.cap = "Another equation using the \\texttt{extrafont} package", echo = FALSE, out.width = "40%"}
include_graphics("../Figures/extrafont-2-embed.pdf")
```

In the case of figure \@ref(fig:bracExampleExtra), note how `extrafont` has once again used the correct font for the greek symbols, but does not quite replicate the rest of the result from \LaTeX{}, as in figure \ref{bracExample}.

\begin{figure}\label{bracExample}
\centering
$\left(\dfrac{x - \mu}{\sigma}\right)$
\caption{A \LaTeX{} equation with brackets}
\end{figure}

```{r results = FALSE}
# Using the 'tikzDevice' package
library(tikzDevice)
options(tikzDocumentDeclaration = "\\documentclass[12pt]{article}")
packages <- c(getOption("tikzLatexPackages"),
              "\\usepackage{amsmath}\n")
options("tikzLatexPackages" = packages) 

tikz("tikzDevice.tex", standAlone = TRUE)
opar <- par(mar = par()$mar + c(0, 1, 0, 0))
plot(x, y, xlab = "$\\mu$", ylab = "", yaxt = "n", type = "l")
axis(2, las = 1)
mtext("$\\dfrac{\\mu^2}{2}$", side = 2, line = 3, las = 1)
dev.off()
```

```{r echo = FALSE, results = "asis"}
tikzTex <- readLines("tikzDevice.tex")
start <- which(grepl("begin\\{document}", tikzTex))
end <- which(grepl("end\\{document}", tikzTex))

beginTikzPic <- which(grepl("begin\\{tikzpicture}", tikzTex))
tikzTex[beginTikzPic] <- gsub("]", ",scale=0.5]", tikzTex[beginTikzPic])
cat(c("\\begin{figure}\\label{tikzDevicePlot}\n\\centering\n", 
      paste0(tikzTex[(start + 1):(end - 1)], 
             collapse = "\n"),
      "\\caption{Drawing the plot with the \\texttt{tikzDevice} package}\n",
      "\\end{figure}"))
```

The `tikzDevice` package in R will draw the axis titles with \TeX{} but will also use \TeX{} for all other text on the plot, like the axis labels in figure \ref{tikzDevicePlot}. This may be appropriate if the graphic is to be included in a \TeX{} document but not so much for other document formats like HTML.

What we want is a middle ground - being able to harness the power of \TeX{} and its typesetting capabilities on *our* choice of text or equation in R graphics. This is where the `dvir` package comes in - providing a simple user interface, in the style of R `grid` graphics, by way of the `grid.latex()` function. Figure \@ref(fig:plotGridLatex) recreates figure \@ref(fig:expressionPlot), but with the axis titles made using `grid.latex()`.

```{r plotGridLatex, fig.cap = "Changing axis titles with \\texttt{grid.latex()}", fig.asp = 1}
library(dvir)
preamble <- "\\documentclass[12pt]{standalone}\n\\usepackage{amsmath}\n\\begin{document}"
par(mar = par()$mar + c(0, 1, 0, 0))
plot(x, y, xlab = "", ylab = "", yaxt = "n", type = "l")
axis(2, las = 1)

grid.latex("$\\mu$", x = 0.545, y = 0.07)
grid.latex("$\\dfrac{\\mu^2}{2}$", x = 0.06, preamble = preamble)
```

## Where this project fits in

The `dvir` package already worked really well in a lot of cases. There were however a number of desirable features of \TeX{} and its extensions that had not yet been implemented by `dvir`. The power of this package comes from ensuring it is comprehensive enough to meet a user's entire \TeX{} needs in R graphics without having to leave R to do annotations in \LaTeX{} itself (or other software like Adobe Photoshop or Illustrator). 

By keeping things "in R" users only need to learn R (and basic \TeX) code to create their graphics and their work is in one place and easily reproducible. It may not be realistic to *completely* replicate \TeX{} in R, however several aspects of the package were identified as having the potential to increase its usefulness. The aspects identified were:

* the speed of the package - anecdotally it took a while to generate graphics, especially if there were many `grid.latex()` calls
* expand `dvir`'s capability of creating \Tikz{} drawings by adding support for linear gradient fills
* adding the ability to align text from `grid.latex()` to a baseline - the natural line on which characters sit