\newpage{}

# Background

## \TeX{}

\TeX{} is a program to format and typeset text, and includes some basic macros to do this. \LaTeX{} is a higher-level implementation that sits on top of \TeX{}, basically consisting of a lot more macros, creating a much more user-friendly interface to \TeX{}. For example, \LaTeX{} allows one to create a document with numbered sections, title pages and bibliographies without having to write complicated \TeX{} macros oneself.

## DVI

A \TeX{} or \LaTeX{} file is just plain text so there needs to be a step to translate this plain text to what you will see on a formatted document on a screen or page. \TeX{} "engine" like `pdflatex` normally processes \TeX{} to PDF but it can also produce a DVI (DeVice Independent) file. A DVI file is a binary file describing the layout of the document. For example, the height of the page, what characters to display and where, and the fonts to be used. DVI provides a format that can be read into R so that we can draw the \TeX{} layout in R. This is what the `dvir` package does.

## The (pre-existing) `dvir` package {#dvirDesc}

In a simplified form, `dvir` works by providing a high level function, `grid.latex()`, to call with the \TeX{} code of the expression or text to be displayed. In this instance, the preamble is changed only because the axis title uses `\dfrac` so the `amsmath` package needs to be specified in the preamble.

```{r, fig.width = 0.4, fig.height = 0.4, fig.cap = "Using the \\texttt{dvir} function \\texttt{grid.latex()}"}
library(dvir)
texCode <- "$\\dfrac{\\mu^2}{2}$"
newPreamble <- 
  "\\documentclass[12pt]{standalone}\n\\usepackage{amsmath}\n\\begin{document}"
grid.latex(texCode, preamble = newPreamble)
```

The following steps are taken when `grid.latex()` runs:

\begin{enumerate}
  \item A \TeX{} document is created with the expression and a changeable default preamble and postamble.
\end{enumerate}

```{r, results = FALSE}
postamble <- paste0("\\end{document}")
cat(paste(newPreamble,
          texCode,
          postamble,
          sep = "\n"))
```

\begin{spacing}{1}
\begin{itemize}[noitemsep]
  \item[] \begin{verbatim}\documentclass[12pt]{standalone}\end{verbatim}
  \item[] \begin{verbatim}\usepackage{amsmath}\end{verbatim}
  \item[] \begin{verbatim}\begin{document}\end{verbatim} 
  \item[] \begin{verbatim}$\dfrac{\mu^2}{2}$\end{verbatim}
  \item[] \begin{verbatim}\end{document}\end{verbatim}
\end{itemize}
\end{spacing}

\begin{enumerate}
  \setcounter{enumi}{1}
  \item This TeX document is then processed using the local \TeX{} installation to create a DVI (DeVice Independent) file.
  \item The DVI file is read into R. As DVI files are binary they are not easily readable by humans but the \texttt{dvir} function \texttt{readDVI()} translates the DVI file into readable text. This is an extract of a DVI file after using \texttt{readDVI()}, stating the definition of a font and then ``setting'' (i.e. printing) a character ``2''.
\end{enumerate}

\begin{spacing}{1}
\begin{itemize}[noitemsep]
  \item[] \begin{verbatim}fnt_def_1    fontnum=15, checksum=2088458503, scale=524288,\end{verbatim}
  \item[] \begin{verbatim}             design=524288, fontname=cmr8\end{verbatim}
  \item[] \begin{verbatim}set_char_50  '2'\end{verbatim}
\end{itemize}
\end{spacing}

\begin{enumerate}
  \setcounter{enumi}{3}
  \item Three ``sweeps'' of the DVI file are completed to extract necessary information about what to display in R (and where and how to display it):
  \begin{itemize}
    \item Font sweep: Gather the names of all fonts used in the DVI file and locate the relevant font files on the local machine. The font information is stored in a R list as well as a \texttt{fontconfig} file.
    \item Metric sweep: To determine the overall bounding box (size) of the expression to display. This bounding box is used to create a \texttt{grid} viewport which can encompass the entire \TeX{} passed to \texttt{grid.latex()} expression using the native DVI coordinates.
    \item Grid sweep: Convert all text and symbols into \textit{grobs} (grid graphical objects) 
  \end{itemize}
  \item These grobs are then displayed in the R graphics device as per the \texttt{grid} package.
\end{enumerate}