\newpage{}

# Code speed (part 2) - font caching

In the previous section, the speed of `dvir` was increased by stopping it doing something "silly". The profiling results from the previous section demonstrate how much time `grid.latex()` spends dealing with fonts so this section is about making `dvir` "smarter" by caching the information it gathers about fonts between `grid.latex()` calls.

For example, creating figure \@ref(fig:yeeExample) requires three `grid.latex()` calls for the same symbol $\omega$. Each time, the following font definition appears in the DVI file for `grid.latex()` to search for during its font sweep.

\begin{spacing}{1}
\begin{verbatim}
fnt_def_1    fontnum=17, checksum=-1209964637, scale=786432, design=786432,
             fontname=cmmi12
\end{verbatim}
\end{spacing}

In fact, that example's 11 `grid.latex()` calls define fonts 29 times in total, for eight unique fonts.

Caching fonts requires several steps:

\begin{itemize}
  \item During the font sweep, search all font definitions as per usual.
  \item For each font definition, check if the font has been cached already.
  \begin{itemize}
    \item If yes, do nothing. Refer to the existing font information in the cache when the needs to be used
    \item If no, search for and store the font information as per usual.
  \end{itemize}
\end{itemize}

The font information is stored in R as a list. Every time `grid.latex()` processed a "preamble" op code in a DVI file, that is, every time `grid.latex()` was called, the font list was initialised as an empty list in the `dvir` namespace with `dvir::set("fonts", vector("list", 255))`. To stop this initialisation happening by default, the following code was put in its place.

```{r, eval = FALSE}
if (dvir::get("initFonts") || is.null(dvir::get("fonts"))) 
  dvir::set("fonts", vector("list", 255))
```

Note in this new statement the logical variable `initFonts`, stored in the `dvir` namespace. This lets a user choose whether to intialise the font cache on a `grid.latex()` call. If `initFonts` is `FALSE` or the font list does not exist, for example after the package is loaded, the font list is initialised. Otherwise, the existing font cache remains. `initFonts` was added as an argument to `grid.latex()` with its default value being equal to the global option `dvir.initFonts`.

In a DVI file, each font definition gives the font a identification number, for example `fontnum=17`. In all the DVI files examined as part of this project, this font identification number is the same when that particular font is used again in another `grid.latex()` call. This number is used as the index of the fonts list where its information is stored. 

When the font sweep passes a font definition in the DVI file it checks first whether the index of the font list matching the font number of the new font contains any existing font information: 

\begin{itemize}
  \item If no, then the font information is written to the font list as usual.
  \item If yes, then it checks if the existing font information is the same as the font information in the DVI file. 
  \begin{itemize}
    \item If yes, do nothing. 
    \item If no, overwrite the font with the new font information.
  \end{itemize}
\end{itemize}

The result of this is that across multiple `grid.latex()` calls requiring the same fonts, any future uses of that font are much faster after the font has been used once. Importantly it also has a "fail safe" where if there happens to be a case where the font number was used with a different font in a previous `grid.latex()` call, the font cache is overwritten with the new font so the correct font will be used.

The updated font list is stored in the `dvir` environment to be retrieved the next `grid.latex()` finds a font definition during a font sweep.

The only weakness of this process is if in a *single* DVI file different fonts are defined with the same font number. In this case the second font will be stored in the font cache in that index and so will be used in place of the first font. No instances of this occurring were discovered during this project.

To examine the effect of these changes on code speed we performed profiling as was detailed in the previous section. For the simple example the time for the font sweep has reduced by 71%, with `grid.latex()` overall taking 45% less time. Similar results can be seen in for the example in figure \@ref(fig:yeeExample) - a 76% and 46% decrease in the font sweep and `grid.latex()` respectively.

```{r echo = FALSE, eval = TRUE}
results2 <- data.frame(time = c(results[[3]][[2]],
                                results[[7]][[2]],
                                results[[3]][[5]],
                                results[[7]][[5]],
                                results[[4]][[2]],
                                results[[8]][[2]],
                                results[[4]][[5]],
                                results[[8]][[5]]),
                       version = rep(c("Before", "After"), 
                                     each = 4),
                       func = rep(c("'gridlatex'", "font"),
                                  each = 2),
                       example = c("simple", "yee"))
results2$time <- results2$time / 20
versionLabels2 <- c(Before = "0.3-1 (before changes)", 
                    After = "0.3-2 (after changes)")
levels2 <- c("gridlatex", "font")
```

```{r speedUp2Simple, fig.cap = "The average time spent in the \\texttt{grid.latex()} function and font sweep, over 20 runs of the simple example.", echo = FALSE, fig.width = 5, fig.height = 4}
plotProfiling1("simple", versionLabels2, results = results2, levels = levels2)
```

```{r speedUp2Yee, fig.cap = "The average time spent in the \\texttt{grid.latex()} function and font sweep, over 20 runs of the complicated example", echo = FALSE, fig.width = 5, fig.height = 4}
plotProfiling1("yee", versionLabels2, results = results2, levels = levels2)
```

## Profiling environment specifications

The exact results obtained in this and the previous section are specific to the computing environment used. Specific details are provided below. The sampling nature of profiling (intermittent recording of the call stack) will give different results every time it is done.

The profiling results are dependent on the computer setup used and could change depending on the exact computing environment in which the `dvir` package is used. 

The profiling results in this report, in this and the previous section, were calculated with the following setup:

- A virtual machine via Oracle VM Virtualbox
- Virtual machine running Ubuntu 18.04.5 LTS
- R version 3.4.4
- `dvir` package versions as described with the profiling results