\newpage{}

<!---
For write up,

Worry about the commas later (worst comes to worst we don't do them)
Replicate for colours
Space in front of stop (see SVG dvi) causes the spaces (also not important, I'd say less important than the commas as it's still readiable by R)

Say added "linear gradient defined" and "linear gradient used" for clarification. Check how these work? Are they just defined in my vert shading section? And called somewhere else? I think so, because PGF doc says it creates a macro


paste(hexView::blockValue(dviLinGradNew[[43]]$blocks$op.opparams.string), collapse = "")


What are those 6 "transform" things in SVG DVI output? I think they are mentioned in dvir package code?

Try making sphere with linear gradient with TikZ, see if fill is still rectangle with clipping?

Try making vertical gradient? is it same output in SVG?

Try making rectangle rotated slightly?

Try rectangle with fill with say, dotted outline (to see if gradioent dill is actuall separate rectangle)
--->

# Linear gradient fills

## \Tikz{} and `dvir`

\Tikz{} [@man-tikz] is a \TeX{} package that allows drawing of pictures and diagrams in \TeX{} documents.

\begin{spacing}{1}
\begin{verbatim}
\begin{tikzpicture}
\path (0, 1) node[circle,fill=red!40,draw,thick] (x) {$\mu$}
      (3, 0) node[circle,fill=blue!40,draw,thick] (y) {$\dfrac{\mu^2}{2}$};
\draw[->] (x) .. controls (1, 0.5) and (2, 0.5).. (y);
\end{tikzpicture}
\end{verbatim}
\end{spacing}

\begin{figure}\label{tikzRaw}
\centering
\begin{tikzpicture}
\path (0, 1) node[circle,fill=red!40,draw,thick] (x) {$\mu$} 
      (3, 0) node[circle,fill=blue!40,draw,thick] (y) {$\dfrac{\mu^2}{2}$};
\draw[->] (x) .. controls (1, 0.5) and (2, 0.5).. (y);
\end{tikzpicture}
\caption{A picture made with \Tikz{}}
\end{figure}

The original DVI specification only needed to account for text, typesetting and the most basic of rectangles and so was not designed with drawing and graphics in mind. The type of instruction in the DVI file are labelled with an "op code". Each op code described a type of instruction like defining fonts, setting characters to display and vertical and horizontal cursor movements. There were four op codes however, called *DVI specials*, that can contain almost any form of instruction or values needed, such as text colour, to create a document based on the DVI file, such as Postscript or PDF.

The \Tikz{} package uses these DVI specials to describe shapes, drawings and colours in PGF (portable graphics format) which can be translated to instructions for other viewing formats, like Postscript, PDF or SVG. How the instructions are translated is controlled by a \Tikz{} driver. The `dvir` package [@murrell-tikz] includes its own \Tikz{} driver to translate the drawing instructions into a form useful to draw the things with R grid graphics [reference Paul dvir \Tikz{} report].

Some \Tikz{} features were not implemented though, notably the ability to have fill colours of shapes as linear or radial gradients or patterns. The primary reason for this is that R did not support these types of fills but the latest R release in May 2021, version 4.1.0, provides support for these fills in the `grid` package [@murrell-grad-fills], on which `dvir` is built.

```{r, include = FALSE}
str2lang <- getFromNamespace("str2lang", "backports")
```

```{r gridTikz, fig.cap = "Using \\texttt{grid.tikzpicture()} to make the diagram from earlier"}
tikzCode <- paste("\\path (0, 1) node[circle,fill=red!40,draw,thick] (x) {$\\mu$}",
                  "      (3, 0) node[circle,fill=blue!40,draw,thick] (y) {$\\dfrac{\\mu^2}{2}$};",
                  "\\draw[->] (x) .. controls (1, 0.5) and (2, 0.5).. (y);",
                  sep = "\n")
tikzPreamble <- gsub("\\\\usepackage\\{tikz}", 
                     "\\\\usepackage{tikz}\n\\\\usepackage{amsmath}", 
                     tikzpicturePreamble())
grid.tikzpicture(tikzCode, preamble = tikzPreamble)
```

The use of `grid.tikzpicture()` is demonstrated in figure \@ref(fig:gridTikz), recreating the diagram from figure \ref{tikzRaw}. The alteration to the preamble once again only necessary due to the use of `\dfrac` from the `amsmath` package.

- [**Example:** \Tikz{} radial gradient fill example]

- [**Example:** Make same \Tikz{} example as above in R with dvir (obviously fill will be blank)]

As it is, the \Tikz{} driver simply ignores any gradient or pattern fill information when creating the DVI file for `dvir`.

- [**Example:** Use `grid.tikzpicture()` for picture with gradient fill in text, but resulting R graphic does not have fill]

## Implementing \Tikz{} linear gradient fills in `dvir`\label{tikzWorkflow}

The following steps are required to implement these \Tikz{} fills in `dvir`:

1. Add the fill information (like the gradient colours and their locations) to the DVI file created by `dvir`

2. Store this fill information during a parse by `dvir` to read the DVI file

3. Add the fill information when drawing the shape in R

To tackle step 1, we need to update the `dvir` \Tikz{} driver file to include information about the gradient and pattern fills. As the `dvir` \Tikz{} driver file is based on the SVG \Tikz{} driver file, the SVG support for \Tikz{} fills was used as a base to edit to make it specific to `dvir`.

<!---
Assuming from here on that we only care about linear gradients

Can I just talk about linear gradient fills here? The work could easily be extended (but wasn't) to radial gradients as well
--->

The information we require for the gradient fills from \Tikz{} via the DVI file is as per the arguments for the `grid::linearGradient(...)`, which is used as an argument to `grid::gpar(fill = linearGradient(...))`, which itself is an argument to a `grid` drawing function, for example `grid::grid.rect(..., gp = gpar(fill = linearGradient(...)))`. The most important parts of defining a linear gradient fill is the colours and stops of the gradient fill. The stops of a gradient fill are the locations along the length of a gradient fill where the specified colours are. In between the stops, the gradient between stop colours either side occurs.

The `colours` and `stops` arguments of `linearGradient()` are simply vectors of colours (a character vector of colour names of hexadecimal RGB values) and locations of those colours as a proportion of the distance between the start and end points of the gradient respectively. This obviously guides us as to what information we need to get from \Tikz{} in the DVI file so we can pass it to `dvir`.

Let us consider a simple example, a rectangle with an orange to green linear gradient fill:
<!---

OLD CODE FOR SOME REASON:

```{eval=FALSE}
# Code from TeX should we want to know how to do this in TeX itself or reduce/simplify to print above the picture below. Could edit to be programatic (hide code chunk to define tikz picture code, then add that in here with results='asis' or generation of picture below)
\documentclass{standalone}
\usepackage{tikz}
\begin{document}
\begin{tikzpicture}
\filldraw [draw=black, left color=orange, right color=green] (0,0) rectangle (4,2);
\end{tikzpicture}
\end{document}
```

```{r results='asis', echo=FALSE}
latexRect <- c("\\begin{center}",
               "\\begin{tikzpicture}",
               "\\filldraw [draw=black, left color=orange, right color=green] (0,0) rectangle (4,2);",
               "\\end{tikzpicture}",
               "\\end{center}")

# y <- paste(latexRect, collapse = "")
# cat(y)
cat(latexRect)
```
--->

\begin{figure}
\centering
\begin{tikzpicture}
\filldraw [draw=black, left color=orange, right color=green] (0,0) rectangle (4,2);
\end{tikzpicture}
\end{figure}

The following is an extract of the DVI file when the rectangle above is generated using the SVG DVI driver included with the common \TeX{} distributions, `pgfsys-dvisvgm.def`. It has been edited slightly for readability by adding line breaks and removing a line break "character", `{?nl}`.
<!---
The output of the linear gradient info below was element 52 after reading in the DVI file into R using 'readDVI()', in case we want to automate this later

CHECK THIS OUTPUT FOR THE 'matrix()' FUNCTION. WHERE DID IT COME FROM? - I think it's okay (check DVI output....)

Maybe remove some spaces if want it this code to go into left margin, or think of where else I could line breaks
--->

```{ dviGradientOutput, eval=FALSE}
xxx1         k=67
             x=dvisvgm:raw <g transform="matrix(1,0,0,1,56.90549,28.45274)"> 
xxx1         k=67
             x=dvisvgm:raw <g transform="matrix(2.26802,0,0,1.134,0.0,0.0)">
xxx1         k=66
             x=dvisvgm:raw <g transform="matrix(0.0,1.0,-1.0,0.0,0.0,0.0)">
xxx4         k=425
             x=dvisvgm:raw  <linearGradient id="pgfsh2" gradientTransform="rotate(90)"> 
                            <stop offset=" 0.0" stop-color=" rgb(0.0%,100.0%,0.0%) "/> 
                            <stop offset=" 0.25" stop-color=" rgb(0.0%,100.0%,0.0%) "/>
                            <stop offset=" 0.5" stop-color=" rgb(50.0%,75.0%,0.0%) "/>
                            <stop offset=" 0.75" stop-color=" rgb(100.0%,50.0%,0.0%) "/>
                            <stop offset=" 1.0" stop-color=" rgb(100.0%,50.0%,0.0%) "/>
                            </linearGradient>
xxx1         k=57
             x=dvisvgm:raw <g transform="translate(-50.1875,-50.1875)"> 
xxx1         k=97
             x=dvisvgm:raw <rect width="100.375" height="100.375" 
                            style="fill:url(#pgfsh2); stroke:none"/> 
```

We can see from this that the linear gradient definition with stops and colours is defined within a `<linearGradient>` element and given an `id` attribute. In the `<rect>` element a CSS style definition sets the fill of the rectangle by referring to the `id` of the previously defined definition. In the linear gradient definition there are colours defined as RGB values and their respective stops so now we need to get the `dvir` driver file to extract the same information in a "R-friendly" form. There are several \TeX{} macros that were defined to do this. These are based on the `pgfsys-common-svg.def` and `pgfsys-dvisgm.def` SVG drivers that come with most \TeX{} distributions.

<!---
- Go to `dvir` driver file and see what I've added (and try work out what bit does what to explain it! Like it gets defined in one place, then passed to another function to do something else)
- See my meeting progress notes on the matter


Test I've got all the bits by copying over to a fresh copy of dvir driver. If it still works, then I've got everything!

Say main difference between SVG and R for stops is that we want to gather stops all together, and their colours all together for R. SVG wants one stops and it's colour togther!
--->

A "wrapper" macro of what to do when a gradient fill is requested. Within this, the definition of the fill is created and sent to the DVI file, followed by a rectangle with a fill specified by that definition. For clarity, a line stating when the gradient fill is defined and then again when it is used has been added to the DVI file but these can be removed.

```
\def\pgfsys@shadinginsidepgfpicture#1{%
  #1%
  \pgfsysprotocol@literal{SHADING BEING DEFINED: 
                          ShadDefID = \the\pgf@sys@dvir@objectcount}%
  \pgf@sys@dvir@sh@defs% 
  \pgf@process{\pgf@sys@dvir@pos}%
  \pgf@xa=-.5\pgf@x%
  \pgf@ya=-.5\pgf@y%
  \pgfsysprotocol@literal{
      <g transform="translate(\pgf@sys@tonumber{\pgf@xa},\pgf@sys@tonumber{\pgf@ya})">}%
  \pgfsysprotocol@literal{SHADING BEING USED: ShadDefID = \the\pgf@sys@dvir@objectcount}
  \pgf@sys@dvir@sh%
}
```

In basic cases, a horizontal linear gradient is defined as a vertical gradient with a 90 degree rotation which is why this is called "vert" shading, as it is in the SCG driver file. The definition and use of the gradient specified above are collated here. The stop positions (proportion along the length of the gradient for which a colour specified) and their respective colours are collated and the literal text to build the gradient definition and rectangle using that definition. The biggest difference from the SVG driver was that R needs a vector of stops and a separate vector of colours as arguments to `linearGradient()`. SVG specifies the position and colour of each stop together.

```
\def\pgfsys@vertshading#1#2#3{%
  {%
    \pgf@parsefunc{#3}%
    \global\advance\pgf@sys@dvir@objectcount by1\relax%
    \pgf@sys@dvir@shading@stops%
    \pgf@sys@dvir@shading@stopcolours%
    \expandafter\xdef\csname @pgfshading#1!\endcsname{%
      \def\noexpand\pgf@sys@dvir@sh@defs{
          \noexpand\pgfsysprotocol@literal{\pgf@sys@dvir@thestops}}%
      \def\noexpand\pgf@sys@dvir@sh{\noexpand\pgfsysprotocol@literal{<rect
        width="\pgf@sys@tonumber{\pgf@y}"
        height="\pgf@sys@tonumber{\pgf@x}"
        style="fill:url(\noexpand\#pgfsh\the\pgf@sys@dvir@objectcount);
          stroke:none"/>\noexpand\pgfsys@dvir@newline}}%
      \def\noexpand\pgf@sys@dvir@pos{\noexpand\pgfpoint{\the\pgf@y}{\the\pgf@x}}%
    }%
  }%
}
```

This macro allows us to iteratively add stop positions or colours to the ones we have already gathered.

```
\let\pgf@sys@dvir@thestops=\pgfutil@empty
\def\pgf@sys@dvir@addtostops#1{%
  \edef\pgf@temp{#1}%
  \expandafter\expandafter\expandafter\def
  \expandafter\expandafter\expandafter\pgf@sys@dvir@thestops
  \expandafter\expandafter\expandafter{
      \expandafter\pgf@sys@dvir@thestops\expandafter\space\pgf@temp}%
}
```

The following macros process all the stop locations, collating them into an R friendly vector suitable to parse to `linearGradient()`

```
\def\pgf@sys@dvir@shading@stops{%
  % Step 1: Compute 1/\pgf@sys@shading@end@pos
  \pgf@x=\pgf@sys@shading@end@pos\relax%
  \c@pgf@counta=\pgf@x\relax%
  \divide\c@pgf@counta by4096\relax%
  % Step 2: Insert stops locations 
  \pgf@sys@dvir@addtostops{stops=(}%
  \expandafter\pgf@sys@dvir@shading@dostoplocations\pgf@sys@shading@ranges%
  % dummy for end:
  {{\pgf@sys@shading@end@pos}{\pgf@sys@shading@end@pos}}%
  \pgf@sys@dvir@addtostops{)}%
}

\def\pgf@sys@dvir@shading@dostoplocations#1{%
  \edef\pgf@test{#1}%
  \ifx\pgf@test\pgfutil@empty%
  \else%
    \expandafter\pgf@sys@dvir@shading@dostoplocation\pgf@test%
    \expandafter\pgf@sys@dvir@shading@dostoplocations
  \fi%
}

\def\pgf@sys@dvir@shading@dostoplocation#1#2#3#4{%
  % #1 start pos
  % #2 end pos
  % #3 start rgb
  % #4 end rgb
  \pgf@x=#1%
  \pgf@x=16\pgf@x%
  \divide\pgf@x by \c@pgf@counta\relax%
  \expandafter\pgf@sys@dvir@addtostops{\pgf@sys@tonumber\pgf@x}%
}
```

Similar to the above, these collate all the colours of the stops into an R friendly vector.

```
\def\pgf@sys@dvir@shading@stopcolours{%
  % Step 1: Compute 1/\pgf@sys@shading@end@pos
  \pgf@x=\pgf@sys@shading@end@pos\relax%
  \c@pgf@counta=\pgf@x\relax%
  \divide\c@pgf@counta by4096\relax%
  % Step 2: Insert stops RGB colours
  \pgf@sys@dvir@addtostops{, colours=(}%
  \expandafter\pgf@sys@dvir@shading@dostopcolours\pgf@sys@shading@ranges%
  % dummy for end:
  {{\pgf@sys@shading@end@rgb}{\pgf@sys@shading@end@rgb}{\pgf@sys@shading@end@rgb}}%
  \pgf@sys@dvir@addtostops{)}%
}

\def\pgf@sys@dvir@shading@dostopcolours#1{%
  \edef\pgf@test{#1}%
  \ifx\pgf@test\pgfutil@empty%
  \else%
    \expandafter\pgf@sys@dvir@shading@dostopcolour\pgf@test%
    \expandafter\pgf@sys@dvir@shading@dostopcolours%
  \fi%
}

\def\pgf@sys@dvir@shading@dostopcolour#1#2#3#4{%
  % #1 start pos
  % #2 end pos
  % #3 start rgb
  % #4 end rgb
  \expandafter\pgf@sys@dvir@shading@dorgb#3%
}
```

This is a helper function to return formatted RGB values.

```
\def\pgf@sys@dvir@shading@dorgb#1#2#3{%
  \pgf@sys@dvir@color@rgb#1,#2,#3\relax%
  \pgf@sys@dvir@addtostops{\pgf@sys@dvir@prepared}%
}
```

A counter is defined in order to give each gradient definition unique identifier.

```
\newcount\pgf@sys@dvir@objectcount
```

The result of these changes mean that the linear gradient definition is specified in the DVI file in a form usable by R, sans some commas:

```
xxx1         k=118
             x=dvir::  stops=( 0.0 0.25 0.5 0.75 1.0 ) , 
                       colours=( rgb(0,1,0) rgb(0,1,0) rgb(0.5,0.75,0) rgb(1,0.5,0) 
                                 rgb(1,0.5,0) )
```

Compare this with the linear gradient definition from earlier:

```
xxx4         k=425
             x=dvisvgm:raw  <linearGradient id="pgfsh2" gradientTransform="rotate(90)">
                            <stop offset=" 0.0" stop-color=" rgb(0.0%,100.0%,0.0%) "/>
                            <stop offset=" 0.25" stop-color=" rgb(0.0%,100.0%,0.0%) "/>
                            <stop offset=" 0.5" stop-color=" rgb(50.0%,75.0%,0.0%) "/>
                            <stop offset=" 0.75" stop-color=" rgb(100.0%,50.0%,0.0%) "/>
                            <stop offset=" 1.0" stop-color=" rgb(100.0%,50.0%,0.0%) "/>
                            </linearGradient>
```

Unfortunately this was the extent this project could explore implementing \Tikz{} gradient fills in `dvir`. You may notice in both the SVG and `dvir` DVI output that the colour at stops 0 and 0.25 are the same, as is the colour at stops 0.75 and 1. This is related to a larger issue of \Tikz{} using a number of transformations on the "simple" linear gradient definition above which is then clipped to the shape the fill is for. These transformations would require some very detailed work to translate them into a set of instructions to perform the same in R before they are used in `linearGradient()`. 

An additional complication is that \Tikz{} will create shapes with these fills by first drawing the shape with a transparent fill and then drawing another shape with the gradient fill clipped to the size of the first shape. R however requires the the fill information as an argument when drawing the shape. In R, any transformations of the gradient, as \Tikz{} does, will need to be done *before* it is parsed to the drawing function.

Radial gradient fills have similar problems with how they are manipulated by \Tikz{}, however the work done so far on linear gradients could be easily replicated for them.

This work is still within step 1 of the workflow in section \ref{tikzWorkflow}. To continue with the implementation, aside from the above obstacles, `dvir` would need to do another sweep of the DVI file, like the font sweep, to achieve step 2. After any transformations or manipulations are made, the gradient definition can be called by a unique identifier to parse the stored fill definitions to the drawing function used as in step 3.

<!---


Section to be continued with:

- What have we had to change in driver file (and why?) - like specific bits of driver file. 

- Before and after of DVI file (using new driver) for linear gradient fill (see new information display). Do the "after" for both simple rectangle and maybe a more complicated example?

- Why couldn't we go further? Mention the transformations?

- Next steps (to complete steps 1, 2 and 3 as detailed earlier), including discussing how this applies to radial gradient fills and pattern fills


Mention how there's the transformastions and clipping, and we somehow need to translate that into a fill for thre shape in R

Remember to talk about the future steps, we need to do a parse of the gradient deifnitions and store them somewhere, then next parse for actual drawing we have to match up the id values

Does SVG do it by actually drawing two rectangles? Or what?

Do I need to explain SVG format at all?

Do we even need to save linear gradient definition? Can it be used in different places, or is it always defined again if the exact same definition is used?

Need to be careful about line length in example of DVI output

maybe end by saying 'linear gradients are just a small step. For example patterns require creating grobs of the pattern definition' etc. etc.

Is this complicated nature of manipulating/transfroming the gradient rectangle also for other formats? or just SVG? I may be asked this in my talk

Currently ignoring pattern fills


--->