\newpage{}

<!---
For write up,

Worry about the commas later (worst comes to worst we don't do them)
Replicate for colours
Space in front of stop (see SVG dvi) causes the spaces (also not important, I'd say less important than the commas as it's still readiable by R)

Say added "linear gradient defined" and "linear gradient used" for clarification. Check how these work? Are they just defined in my vert shading section? And called somewhere else? I think so, because PGF doc says it creates a macro


paste(hexView::blockValue(dviLinGradNew[[43]]$blocks$op.opparams.string), collapse = "")



--->

# Linear gradient fills

## TikZ and `dvir`

TikZ is a \TeX{} package that allows drawing of pictures and diagrams in \TeX{} documents [reference TikZ report/description]:

- [**Example:** simple TikZ drawing (circles with labels, and an arrow maybe)]

- [**Example:** more complicated TikZ drawing, maybe with colouring and stuff]

The original DVI specification only needed to account for text and typesetting (and can do the most basic of rectangles too!), and so was not designed with drawing and graphics in mind. The type of instruction in the DVI file are labelled with an "op code". Each op code described a type of instruction like defining fonts, setting characters to display and vertical and horizontal cursor movements. There were four op codes however, called *DVI specials*, that can contain almost any form of instruction or values needed, such as text colour, to create a document based on the DVI file, such as Postscript or PDF.

The TikZ package uses these DVI specials to describe shapes, drawings and colours in PGF (portable graphics format) which can be translated to instructions for other viewing formats, like Postscript, PDF or SVG. How the instructions are translated is controlled by a TikZ driver. The `dvir` package includes its own TikZ driver to translate the drawing instructions into a form useful to draw the things with R grid graphics [reference Paul dvir TikZ report].

Some TikZ features were not implemented though, notably the ability to have fill colours of shapes as linear or radial gradients or patterns. The primary reason for this is that R did not support these types of fills but the latest R release in May 2021, version 4.1.0, provides support for these fills in the `grid` package, on which `dvir` is built.

- [**Example:** replicate one of the above examples in R]

- [**Example:** TikZ radial gradient fill example]

- [**Example:** Make same TikZ example as above in R with dvir (obviously fill will be blank)]

- [**Example:** Use R 4.1.0 to make a linear gradient in a shape]

As it is, the TikZ driver simply ignores any gradient or pattern fill information when creating the DVI file for `dvir`.

- [**Example:** Use `grid.tikzpicture()` for picture with gradient fill in text, but resulting R graphic does not have fill]

## Implementing TikZ linear gradient fills in `dvir`

The following steps are required to implement these TikZ fills in `dvir`:

1. Add the fill information (like gradient start and end colours, gradient radius etc.) to the DVI file created by `dvir`

2. Store this fill information during a parse by `dvir` to read the DVI file

3. Add the fill information when drawing the shape in R

To tackle step 1, we need to update the `dvir` TikZ driver file to include information about the gradient and pattern fills. As the `dvir` TikZ driver file is based on the SVG TikZ driver file, the SVG support for TikZ fills was used as a base to edit to make it specific to `dvir`.

<!---
Assuming from here on that we only care about linear gradients

Can I just talk about linear gradient fills here? The work could easily be extended (but wasn't) to radial gradients as well
--->

The information we require for the gradient fills from TikZ via the DVI file is as per the arguments for the `grid::linearGradient(...)`, which is used as an argument to `grid::gpar(fill = linearGradient(...))`, which itself is an argument to a `grid` drawing function, for example `grid::grid.rect(..., gp = gpar(fill = linearGradient(...)))`. The most important parts of defining a linear gradient fill is the colours and stops of the gradient fill. The stops of a gradient fill are the locations along the length of a gradient fill where the specified colours are. In between the stops, the gradient between stop colours either side occurs.

The `colours` and `stops` arguments of `linearGradient()` are simply vectors of colours (a character vector of colour names of hexadecimal RGB values) and locations of those colours as a proportion of the distance between the start and end points of the gradient respectively. This obviously guides us as to what information we need to get from TikZ in the DVI file so we can pass it to `dvir`.

Let us consider a simple example, a rectangle with an orange to green linear gradient fill:

```{eval=FALSE}
# Code from TeX should we want to know how to do this in TeX itself or reduce/simplify to print above the picture below. Could edit to be programatic (hide code chunk to define tikz picture code, then add that in here with results='asis' or generation of picture below)
\documentclass{standalone}
\usepackage{tikz}
\begin{document}
\begin{tikzpicture}
\filldraw [draw=black, left color=orange, right color=green] (0,0) rectangle (4,2);
\end{tikzpicture}
\end{document}
```

```{r results='asis', echo=FALSE}
latexRect <- c("\\begin{center}",
               "\\begin{tikzpicture}",
               "\\filldraw [draw=black, left color=orange, right color=green] (0,0) rectangle (4,2);",
               "\\end{tikzpicture}",
               "\\end{center}")

# y <- paste(latexRect, collapse = "")
# cat(y)
cat(latexRect)
```

The following is an extract of the DVI file when the rectangle above is generated using the SVG DVI driver included with the common \TeX{} distributions, `pgfsys-dvisvgm.def`. It has been edited slightly for readability.
<!---
The output of the linear gradient info below was element 52 after reading in the DVI file into R using 'readDVI()', in case we want to automate this later

CHECK THIS OUTPUT FOR THE 'matrix()' FUNCTION. WHERE DID IT COME FROM?

Maybe remove some spaces if want it this code to go into left margin, or think of where else I could line breaks
--->

```{r dviGradientOutput, eval=FALSE}
xxx1         k=67
             x=dvisvgm:raw <g transform="matrix(1,0,0,1,56.90549,28.45274)">{?nl} 
xxx1         k=67
             x=dvisvgm:raw <g transform="matrix(2.26802,0,0,1.134,0.0,0.0)">{?nl} 
xxx1         k=66
             x=dvisvgm:raw <g transform="matrix(0.0,1.0,-1.0,0.0,0.0,0.0)">{?nl} 
xxx4         k=425
             x=dvisvgm:raw  <linearGradient id="pgfsh2" gradientTransform="rotate(90)">{?nl} 
                            <stop offset=" 0.0" stop-color=" rgb(0.0%,100.0%,0.0%) "/>{?nl} 
                            <stop offset=" 0.25" stop-color=" rgb(0.0%,100.0%,0.0%) "/>{?nl} 
                            <stop offset=" 0.5" stop-color=" rgb(50.0%,75.0%,0.0%) "/>{?nl} 
                            <stop offset=" 0.75" stop-color=" rgb(100.0%,50.0%,0.0%) "/>{?nl} 
                            <stop offset=" 1.0" stop-color=" rgb(100.0%,50.0%,0.0%) "/>{?nl} 
                            </linearGradient>{?nl} 
xxx1         k=57
             x=dvisvgm:raw <g transform="translate(-50.1875,-50.1875)"> 
xxx1         k=97
             x=dvisvgm:raw <rect width="100.375" height="100.375" style="fill:url(#pgfsh2); 
                            stroke:none"/>{?nl} 
```

We can see from this that the linear gradient definition with stops and colours is defined within a `<linearGradient>` element and given an `id` attribute. In the `<rect>` element a CSS style definition sets the fill of the rectangle by referring to the `id` of the previously defined definition. We can see in the linear gradient definition there are colours defined as RGB values and their respective stops so now we need to get the `dvir` driver file to extract the same information in a "R-friendly" form.

Section to be continued with:
<!---
- Go to `dvir` driver file and see what I've added (and try work out what bit does what to explain it! Like it gets defined in one place, then passed to another function to do something else)
- See my meeting progress notes on the matter
--->

- What have we had to change in driver file (and why?) - like specific bits of driver file

- Before and after of DVI file (using new driver) for linear gradient fill (see new information display). Do the "after" for both simple rectangle and maybe a more complicated example?

- Why couldn't we go further? Mention the transformations?

- Next steps (to complete steps 1, 2 and 3 as detailed earlier), including discussing how this applies to radial gradient fills and pattern fills



<!---
Remember to talk about the future steps, we need to do a parse of the gradient deifnitions and store them somewhere, then next parse for actual drawing we have to match up the id values

Does SVG do it by actually drawing two rectangles? Or what?

Do I need to explain SVG format at all?

Do we even need to save linear gradient definition? Can it be used in different places, or is it always defined again if the exact same definition is used?

Need to be careful about line length in example of DVI output

maybe end by saying 'linear gradients are just a small step. For example patterns require creating grobs of the pattern definition' etc. etc.
--->