\newpage{}

# Text baselines

```{r, include = FALSE}
# Setup the functions etc. for this section
source("../Appendices/algorithms.R")

testText <- c("testing",
              "$\\sum_{n=1}^{\\infty} 2^{-n} = 1$",
              "$\\sum\\limits_{n=1}^{\\infty} 2^{-n} = 1$",
              "The equation is $x + \\frac{\\mu^2}{2}$",
              "$\\frac{\\mu^2}{2} + x$ is the equation",
              "\\begin{minipage}{1in}Multiline, with some line wrapping\\end{minipage}")

# Function to convert 'grid' units to LaTeX units (changing the characters at the end basically)
unitConvert <- function(x) {
  paste0(as.numeric(x), switch(attr(x, "unit"),
                               scaledpts = "sp",
                               inches = "in",
                               points = "pt",
                               attr(x, "unit")))
}
# Function to print text and do draw baselines for given text and baseline(s)
raiseBox <- function(text, baselineVals) {
  latex <- paste0("\\settodepth{\\dvidepth} {", 
                  text, 
                  "}%\n\\settowidth{\\dviwidth} {",
                  text,
                  "}%\n")
  
  latex <- paste0(latex, 
                  paste0("\\raisebox{\\dimexpr -\\dvidepth+", 
                         unitConvert(baselineVals),
                         "}{\\makebox[0in][l]{\\rule{\\dviwidth}{.1pt}}}%\n", 
                         collapse = ""),
                  text,
                  "\n\n",
                  collapse = "")
  latex
}

# Wrapper to call \raiseBox over multiple examples, with multiple sets of baseline(s)
makeFigure <- function(examples, allBaselineVals, caption = "", label = "") {
  baselineEnviroHead <- paste0("\\begin{figure}\n",
                               "\\begin{center}\n")
  baselineEnviroFoot <- paste0("\\end{center}\n\\caption{",
                               caption,
                               "}\\label{",
                               label,
                               "}\n\\end{figure}\n")
  code <- baselineEnviroHead
  for (s in seq_along(examples)) {
    code <- c(code, 
              raiseBox(examples[s], allBaselineVals[[s]]),
              "\\vspace{0.3cm}\n")
  }
  code <- c(code, baselineEnviroFoot)
  
  paste0(code, collapse = "")
}

# Wrapper to return baseline values for multiple pieces of text using baselines() function
returnBaselines <- function(testText,
                            algo, 
                            dviMovesMethod = "all", 
                            dviMovesSelection = "1") {
  algorithmBaselines <- list()
  for (i in seq_along(testText)) {
      algorithmBaselines[[i]] <- baselines(tex = testText[i],
                                           algorithm = algo,
                                           dviMovesMethod = dviMovesMethod,
                                           dviMovesSelection = dviMovesSelection)
  }
  algorithmBaselines
}
```

\newlength{\dviwidth}

\newlength{\dvidepth}

\setlength{\fboxsep}{0in}

\setlength{\parindent}{0in}

## The problem

Text characters have a baseline, that is, a horizontal line on which the characters naturally sit so all the letters appear to be in line with each other. Some letters, like a lower case y or j, have a ``descender''. A descender is a part of a character that sits below the baseline. 

```{r, echo = FALSE, fig.cap = "The baseline and descender of the letter y", fig.height = 0.8, fig.width = 3}
yAlign <- 0.25
grid.text("y", y = yAlign, gp = gpar(fontsize = 50), just = c("left", "bottom"))

xVal <- 0.48
xOffset <- 0.015
xOffsetDepthLabel <- 0.02
yValBottom <- 0.073
yValTop <- yAlign

grid.lines(x = c(xVal + xOffset, 0.65), y = yAlign, gp = gpar(lty = "dashed"))
grid.lines(x = xVal, y = c(yValBottom, yValTop))
grid.lines(x = xVal + c(xOffset, 0), y = yValBottom)
grid.lines(x = xVal + c(xOffset, 0), y = yValTop)
grid.lines(x = xVal - c(xOffsetDepthLabel, 0), y = (yValTop + yValBottom) / 2)

grid.text("baseline", 
          x = 0.662, 
          y = yAlign - 0.005,
          just = "left", 
          gp = gpar(fontsize = 8, fontface = "italic"))
grid.text("descender", 
          x = xVal - xOffsetDepthLabel - 0.01, 
          y = (yValTop + yValBottom) / 2, 
          just = "right",
          gp = gpar(fontsize = 8, fontface = "italic"))
```

The `grid.text()` function accounts for this baseline so when you bottom-align text at a certain y-value, the descenders will actually fall below the y-value we defined, despite the bottom justification. Unfortunately, `dvir` does not account for text baselines and will do any alignment in relation to bounding box of the text, which, in the case of figure \@ref(fig:baselineCombine), is at the bottom of the descenders.

```{r baselineCombine, fig.cap = "Combining \\texttt{grid.text()} and \\texttt{grid.latex()}", fig.height = 0.7, fig.width = 4.5}
texText <- "{\\LARGE changing to \\LaTeX{}}"
grid.text("An example of ", 
          y = 0.1, 
          just = c("right", "bottom"), 
          gp = gpar(fontsize = 18))
grid.latex(texText, y = 0.1, just = c("left", "bottom"))
grid.lines(x = c(0.1, 1), y = 0.1, gp = gpar(lwd = 0.5))
```

To fix this, `dvir` needs to obtain or calculate a value for the baseline for any given piece of text to use as an offset for the bounding box alignment when it is rendering the text. From this point on, this offset will be referred to as a baseline value. All of `dvir`'s information comes from the DVI file so it either needs to extract a baseline value from the DVI file itself, or calculate it *from* information in the DVI file. Unfortunately plain DVI files do not state a baseline value, rather they only detail specific placement of each character. As such, some possible methods to obtain a baseline value were explored. These methods are referred to as algorithms from here on due to their heuristic nature.

## Implementation

To explore the algorithms detailed below and evaluate their usefulness an R function, `baselines()`, has been created. This function can be found in the appendix in the file `algorithms.R`. This function takes several arguments, including the baseline selection algorithm as detailed below, any additional information needed for that particular algorithm, and the \TeX{} code as you would use with `grid.latex()`. The output of this function is the baseline value or values. These values are returned as `grid` units as they are distances from the bottom of the bounding box of the text to the text's baseline.

Once the baseline values have been calculated the bounding box of the text can be bottom-aligned with the y-value specified but then moved down by the baseline value. This means the baseline of the text, should the algorithms work correctly, will be aligned at the specified y-value.

This function has been written to easily allow integration of other algorithms and most of the function could be directly implemented in the `dvir` package, should this baseline algorithm feature be implemented into `dvir` formally.

## Potential solutions

Several different algorithms were explored to calculate the baseline for different types of text that could be used with `grid.latex()`. These algorithms are detailed here.

### `alex` algorithm

This is a simple algorithm which was determined after inspection of some DVI files. In every DVI file, there is a statement specifying the size of the bounding box of the text.

\begin{spacing}{1}
\begin{verbatim}
xxx1         k=50
             x=ps::%%HiResBoundingBox: 0 0 14.21944pt 25.85977pt.
\end{verbatim}
\end{spacing}

After this statement there appears to consistently be a downward move equal to the height of the bounding box of text (from the top left of the bounding box to the bottom left), and then a move upward before the first character is drawn. This algorithms take the cursor location after that upward move to be the baseline. In instances where the entire text has no descenders (i.e. the baseline is the bottom of the bounding box) there will be no upward move before the first character and the baseline value returned is 0.

```{r baselineDemo, fig.cap = "Using the \\texttt{baselines()} function, returning the baseline of \\texttt{$x - \\mu$} with the \\texttt{alex} algorithm"}
baselineValue <- baselines(tex = "$x - \\mu$", algorithm = "alex")
baselineValue
```

```{r results = "asis", echo = FALSE}
algo <- "alex"
cat(makeFigure(testText, 
               returnBaselines(testText, 
                               algo, 
                               dviMovesMethod = "all", 
                               dviMovesSelection = "1"), 
               "The baselines as calculated by the \\texttt{alex} algorithm",
               label = "alexAlgo"))
```

For plain text this algorithm works quite well. Unfortunately for many equations, particularly ones with superscripts or subscripts, the first upward move is *not* to where the baseline is. Note in figure \ref{alexAlgo} the `alex` algorithm has calculated the baseline of the fifth example to be the bottom of the $\mu$. This demonstrates that this algorithm is only practical when the first character's baseline is the one to be aligned, and in the case of the second example, does not serve much purpose at all.

### `dviMoves` algorithm

This is an extension of the `alex` algorithm. Rather than only taking the location after the second vertical cursor move, this algorithm keeps track of *all* the up and down moves of the cursor. The motivation behind this is that the upward and downward moves in the DVI file should reflect the cursor moving to the baseline value of the next character to be typeset. Once again we assume that the first downward move after the "HiResBoundingBox" statement is from the top left to the bottom left of the bounding box so this algorithm only returns the upward and downward moves from there. As DVI files have the ability to save the current cursor location, move around a bit, then reset back to the saved location, all up and down moves are tracked from the start of the DVI file.

There are two complications with this method:

- As it returns *all* the vertical positions the cursor moves to there are many possible baseline values returned. Any more than one means we have to decide which of them to choose.

- There are often several up and downward moves in the DVI file between typeset characters so in between the "useful" baseline values there can be some which are duplicates or are not very useful.

To account for these considerations, along with the `dviMoves` algorithm, the `baselines()` function also allows a choice of method to select a *single* baseline out of the many returned by the algorithm.

#### `dviMoves` selection method `all`

This selection method will return all the baseline values as determined by the algorithm. This is useful for drawing the text with multiple lines representing all the baseline values calculated from this method.

#### `dviMoves` selection method `index`

When this selection method is chosen, another argument to the function, `dviMovesSelection`, is used to specify a numeric index. The baseline value corresponding to that index is returned.

#### `dviMoves` selection method `bottomUp`

This is similar to the `index` method but the baseline values are first ordered from smallest to largest, before using the `dviMovesSelection` argument to select the index of the baseline value to return.

#### `dviMoves` selection method `nextChars`

This method will return the first baseline value just after a specified character in the argument `dviMovesSelection`. Characters are specified by a number, entered as a character string. This method is limited for several reasons:

* It cannot return a baseline value for any instance of a character after the first instance of that character
* If multiple cursor moves are made before a character is drawn, the earlier moves are still recorded but with an empty character label
* Some characters are hard to type on a "normal" keyboard, for example $\mu$
* The DVI file does not actually contain the character, but rather the numeric index of it in the particular font. This is why a number has to be used to select which character it is.

```{r results = "asis", echo = FALSE}
algo <- "dviMoves"
cat(makeFigure(testText, 
               returnBaselines(testText, 
                               algo, 
                               dviMovesMethod = "all", 
                               dviMovesSelection = "1"), 
               "The baselines as calculated by the \\texttt{dviMoves} algorithm",
               label = "dviMovesAlgo"))
```

This algorithm performs very well. In figure \ref{dviMovesAlgo} nearly every part of all examples has a baseline value associated with it, including the superscripts of math equations and every line of the multiline text. This means alignment would be possible with these parts of the examples.

### `preview` algorithm

This algorithm uses the `preview` package [@man-preview] in \LaTeX{}. @preview-tug describes the use of the algorithm for `dvisvgm`, and it has been implemented here as it has in `matplotlib.texmanager` [@matplotlib]. 

When this algorithm is selected, the following code is added to the header of the DVI file:

\begin{spacing}{1}
\begin{verbatim}
\usepackage[active,showbox,tightpage]{preview}
\def\\showbox#1%%
{\immediate\\write16{MatplotlibBox:(\\the\\ht#1+\\the\\dp#1)x\\the\\wd#1}}
\end{verbatim}
\end{spacing}

This results in the following line being printed in the log file.

\begin{spacing}{1}
\begin{verbatim}
MatplotlibBox:(8.04175pt+3.00005pt)x62.81718pt
\end{verbatim}
\end{spacing}

The baseline value in this case is the value just after the `+` and can be obtained by searching the log file for this line starting with `MatplotlibBox:` and returning the value just after the `+`.

```{r results = "asis", echo = FALSE}
algo <- "preview"
cat(makeFigure(testText, 
               returnBaselines(testText, 
                               algo, 
                               dviMovesMethod = "all", 
                               dviMovesSelection = "1"), 
               "The baselines as calculated by the \\texttt{preview} algorithm",
               label = "previewAlgo"))
```

This algorithm also performs very well. In all examples in figure \ref{previewAlgo} the natural baseline of the text has been identified and so this algorithm would be suitable in most use cases. For the multiline text, the second line of text has been identified as having "the" baseline.

### `dvipng` algorithm

The `dvipng` program [@man-dvipng] turns DVI files into PNG images. An option of `dvipng` is `--depth` which returns the baseline value in pixels. [reference dvipng package - maybe where you egt it from linux?]

The system call `dvipng --depth test.dvi` returns `depth=4` within its output so after recording the output, a simple search for `depth=` will find the baseline value.

```{r results = "asis", echo = FALSE}
algo <- "dvipng"
cat(makeFigure(testText, 
               returnBaselines(testText, 
                               algo, 
                               dviMovesMethod = "all", 
                               dviMovesSelection = "1"), 
               "The baselines as calculated by the \\texttt{dvipng} algorithm",
               label = "dvipngAlgo"))
```

It may not be obvious in printed form in figure \ref{dvipngAlgo} but the only example with which the `dvipng` algorithm has selected a "good" baseline value is one of the equations. The other examples have baseline values which are at least slightly off. This is likely a side effect of this algorithm returning the baseline value in (whole) pixels which does not lend itself well to this detailed application.

## Discussion of these algorithms

Of the algorithms considered, the `dviMoves` algorithm performed best. While these algorithms (except `dvipng`) perform well for some or most of the examples, the `dviMoves` algorithm performs well for *all* the examples. Notably, it gives the option to align the baseline of *any* line of multiline text or with almost any character in a mathematical equation, whether it be of a different size, part of a fraction, or a superscript or subscript.

The difficulty with the algorithm is how to choose a *single* baseline value from all the ones calculated. It is possible to manually check them all the baseline values to find the right one but an automated selection method would be desirable. As the `preview` algorithm calculated baseline values for all examples except the multiline text, the algorithms would work well together - an implementation of this functionality in `dvir` could calculate the baseline values with both the `dviMoves` and `preview` algorithms with the default selection being the baseline value from `dviMoves` that is closest to the baseline value calculated by the `preview` algorithm. This could be overridden to select another baseline value calculated by `dviMoves`.