\newpage{}

<!---
Give examples throughout this section of function use (like what the output looks like for example, if return a unit value), so don't need to do it when talking about the specific algorithms

Would it be possible to go into each character in the font and get the font metrics?

VERY IMPORTANT TO TRY OUT LATEX WITH THESE, IN CASE THEY DON'T WORK

Remember for things like multi line text, would need to add baselines of each line to the total height from the text above it (or from below)



In algorithms.R, delete "plotting" stuff (currently commented out) if we don't want that to be in final product
Maybe could move plotting stuff to "examples" script? Probably nah because of the fact they don't get the right baseline in R

Need to go through algorithms.R and tidy up, and proofread comments and all that


Explicitly mention name of algorithms.R function, and maybe the examples script too?, say they're in appendix
--->

# Text baselines

Section outline:

- Do all this without examples first maybe, as that will be some fiddly work?

- Demonstrate problem (with example using grid.text), especially try multi line text maybe?

- Describe algorithms for determining baselines one by one. In dviMoves, describe then the issues with choosing which one, and the potential algorithms for that

- Describe function I made to calculate all baselines using these methods

- Show result of all this (in LaTeX)

## The problem

Text characters have a baseline, that is, a horizontal line on which the characters naturally sit so all the letters appear to be in line with each other. Some letters, like a lower case p or j, have a "descender". A descender is the part of a character that sits *below* the baseline. `grid.text()` accounts for this baseline so when you bottom align text at a certain y value, the descenders will actually fall below the y value we defined, despite the bottom justification. This makes sense because nearly every piece of written text is like this. Unfortunately, `grid.latex()` doesn't account for text baselines. It will simply do any alignment in relation to bounding box of the text.

To fix this, `dvir` needs to obtain or calculate a value for the baseline for any given piece of text to offset the bounding box when it is drawing text. All of `dvir`'s information comes from the DVI file we either need to get a baseline value from the DVI file itself, or calculate it *from* information in the DVI file. Unfortunately DVI files do not state a baseline value so we will have to explore some possible methods to determine a baseline value. These methods are referred to as algorithms from here on due to their heuristic nature.

## Implementation

To explore the practicality of the algorithms detailed below and evaluate their usefulness an R function, `baselines()` has been created as part of this project. This function takes several arguments, including the desired baseline selection algorithm, any other information needed for that particular algorithm, and the \TeX{} code as you would use with `grid.latex()`. The output of this function is the distance, or in some case distances, from the bottom of the bounding box of the text to the possible baseline value. These distances are return as `grid` units. 

Now that the potential baselines have been calculated, if when displaying the text we do not use the specified y value, but rather move the bounding box down by the amount of the baseline height, it will be the baseline of the text that is equal to the y-value.

This function has been written to easily allow integration of other algorithms and most of the function should be able to be directly implemented in the `dvir` package, should this baseline algorithm feature be implemented into `dvir` formally.

## Our potential solutions

We explored several different algorithms to calculate the baseline for several different types of text that could be used with `grid.latex()`. These algorithms are detailed below.

### `alex` algorithm

This is a simple algorithm which was determined after inspection of some DVI files. In every DVI file, there is a statement specifying the size of the bounding box of the text.

- **Example** of of the HiResBoundingBox statement here 

After this statement there appears to consistently be a downward move the height of the bounding box of text, and then a move upward before the first character is drawn. This algorithms take the cursor location after that upward move to be the baseline. In instances where the entire text has no descenders i.e. the baseline is the bottom of the bounding box, there will be no upward move before the first character. In this case a value of 0 is returned as the baseline.

Here is an example of the `baselineAlgorithms()` function in use...

... and here are some examples showing the text and the calculated baseline from the function...

<!---
Discussion - what does it work well or not well for?
--->

<!---
Where I define testText below - should I either put it in a non-visible chunk to evaluate in abckground, or just call the examples I'm going to do for each example indicudally?
--->

<!---
Here's the LaTeX stuff I need for examples!

https://www.techwalla.com/articles/horizontal-line-options-in-latex

remove eval=TRUE from these code chunks
--->
```{r, eval = TRUE}
x <- grid::unit(5, "cm")
y <- "hello"

```

\newsavebox{\dvibox}

\newcommand{\boxreport}[2]{%
  \sbox{\dvibox}{\parbox[b]{1in}{#1}}%
  \usebox{\dvibox}%
  \vspace{-#2}%
  \hrule%
  \vspace{#2}%
}

\boxreport{testing}{0cm}

\boxreport{testing}{0.5mm}

\newcommand\rText{`r paste0(x)`}

\rText

\newcommand\sText{`r y`}

\sText


```{r, eval = TRUE}
source("../Appendices/algorithms.R")
testText <- c("test",       # Text with no danglies
              "testing",    # Text with danglies
              "var",        # Short text with no danglies
              "varying",    # Short text with danglies
              "$\\sum_{n=1}^{\\infty} 2^{-n} = 1$",             # Equation only (inline)
              "$\\sum\\limits_{n=1}^{\\infty} 2^{-n} = 1$",     # Equation only (inline, but like 'displaymath')
              "$\\sum\\limits_{n=1}^{} 2^{-n} = 1$",            # Equation only (inline, but like 'displaymath' and without upper summation limit)
              "The equation is $x + \\frac{\\mu^2}{2}$",        # Text then equation
              "The equation is $x + \\frac{\\mu}{2}$",          # Text then equation (with lower equation height)
              "$\\frac{\\mu^2}{2} + x$ is the equation",        # Equation then text (with fraction first)
              "$x + \\frac{\\mu}{2}$ is the equation",          # Equation then text (with lower equation height, and fraction first)
              "\\begin{minipage}{1in}Paragraph, with some line wrapping!\\end{minipage}")  # Multiline text/paragraph

alexAlgorithmBaseline <- baselines(tex = testText[5], 
                                   algorithm = "alex")
alexAlgorithmBaseline
```





### `dviMoves` algorithm

This is an extension of the `alex` algorithm. Rather than only taking the location after the second 'down' move, this algorithm keeps track of *all* the up and down moves of the cursor. The motivation behind this is that the upward and downward "moves" in the DVI file reflect the cursor moving to the baseline value of the next character to be typeset. Once again we assume that the first downward move after the "HiResBoundingBox" statement id from the top to the bottom of the bounding box and so this algorithm only returns the upward and downward cursor moves from there, however as DVI files have the ability to save the current cursor location, move around a bit, then reset back to the saved location, all up and down moves are recorded from the start of the DVI file.

There are two complications with this method:

- As it returns all the vertical positions the cursor moves to there are many possible "baselines" returned. Any more than one means we have to decide which of the baselines values to actually choose

- There are often several up and downward moves in the DVI file between typeset characters so in between the "useful" baselines there can be some which are not so useful or duplicates

To account for these considerations, along with the `dviMoves` algorithm, the function also allows a choice of method to select a *single* baseline out of the usually many returned by the algorithm. These methods are:

#### `dviMoves` selection method `all`

<!---
Maybe add example doing length(v), just to show how many potential baselines there are for some examples
--->


#### `dviMoves` selection method `index`



#### `dviMoves` selection method `bottomUp`



#### `dviMoves` selection method `nextChars`



#### Other potential `dviMoves` selection methods (not implemented)

- best guess

- prev char

- extend dviMoves to only record position just before a character is typeset?



<!---
Discussion - what does dviMoves work well or not well for?
--->

### `preview` algorithm




<!---
Discussion - what does it work well or not well for?
--->

### `dvipng` algorithm





<!---
Discussion - what does it work well or not well for?
--->

### any other algorithm?

## Discussion of algorithms

<!---
Need to plot these things with latex to see if DVI moves is actuall best? And if it works for all situations?

--->
Overall, the `dviMoves` algorithm has performed best out of all of these. While most of these algorithms perform well for most of the examples, the `dviMoves` algorithm performs well for *all* the examples, notably for giving the option to align the baseline of *any* line of multi-line text. Utilising the fact it returns all baseline values for all displayed characters, it is possible to align, for example, with any character in a mathematical equation, whether it be of a different size or a superscript or subscript.


## Next steps to integrate with `dvir` package

- mention the "matching' algorithm of dviMoves (to compare with another algorithm)

- mention the "just ask TeX for answer" method
